---
title: "NYPD Incident Report"
author: "Jessica W"
date: "2024-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning= FALSE, message = FALSE)

```

```{r}
# Importing necessary libraries
library(dplyr)
library(tibble)
library(hms)
library(lubridate)
library(leaflet)
library(purrr)
library(ggplot2)
```


Description:

Data contains the shooting incidents that occurred In New York City between 2006 and the end of the previous calender year. The records contain information about the location and time of the each incident as well as information about the suspects and victims of each case

```{r import file}

# Read in csv and print a summary for overview of data
incident_data <- "../R_projects/NYPD_Shooting_Incident_Data__Historic_.csv"
data <- read.csv(incident_data)
summary(data)

```



```{r data cleanup, echo=TRUE, warning=FALSE}

# Group together categorical columns to change to factor
factor_vars <- c('PERP_AGE_GROUP', 'PERP_SEX', 'PERP_RACE', 'VIC_AGE_GROUP',
                 'VIC_SEX', 'VIC_RACE', 'BORO')

cleaned_data <- data %>%
  # Desired columns
  select(all_of(c(factor_vars, 'OCCUR_DATE', 'OCCUR_TIME', 'Latitude', 'Longitude', 'STATISTICAL_MURDER_FLAG'))) %>%
  # Change all data point values where an empty string, (null) or unknown to NA
  # to fix any missing data also to condense group sizes when using summary
  mutate(across(everything(), ~case_when(
    . == "" | . == "(null)" | . == 'UNKNOWN' | . == 'U' ~ NA, TRUE ~ . ))) %>% 
  # Change date, time, and factor datatype
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
         OCCUR_TIME = as_hms(OCCUR_TIME),
         across(all_of(factor_vars), as.factor))

summary(cleaned_data)

```
```{r map_labels}

# Create new column with a label id for creating the colors/legend on the map
labeled_data <- cleaned_data %>%
  filter(!is.na(PERP_RACE)) %>% 
  mutate(race_id = case_when(
    PERP_RACE == 'AMERICAN INDIAN/ALASKAN NATIVE' ~ 1,
    PERP_RACE == 'ASIAN / PACIFIC ISLANDER' ~ 2,
    PERP_RACE == 'BLACK' ~ 3,
    PERP_RACE == 'BLACK HISPANIC' ~ 4,
    PERP_RACE == 'WHITE' ~ 5,
    PERP_RACE == 'WHITE HISPANIC' ~ 6,
  ))

```

```{r map_plot}

# Create base map around NYC
ny_map <- leaflet() %>%
  setView(lng = -74.006, lat = 40.7128, zoom = 10)  # Centered around NYC

# Add default OpenStreetMap tiles
ny_map <- ny_map %>% addTiles()

# Define start and end dates for filtering
start_date <- as.Date("2021-01-01")
end_date <- as.Date("2022-01-31")

# Color palette based off of the race_id column
color_palette <- colorFactor("Dark2", levels = unique(labeled_data$race_id))

# Add circle markers using lat and long from the data
ny_map <- ny_map %>% addCircleMarkers(
  data = labeled_data %>% 
  filter(OCCUR_DATE >= start_date & OCCUR_DATE <= end_date),
  lng = ~Longitude,
  lat = ~Latitude,
  radius = 1,
  color = ~color_palette(race_id),
  group = ~PERP_RACE  # Use category_column for the legend items
  
)

# Add the legend and the ability to filter out the legend items
ny_map <- ny_map %>%
  addLayersControl(
    # Specify unique categories for the legend
    overlayGroups = unique(labeled_data$PERP_RACE),
    # Expand the legend by default
    options = layersControlOptions(collapsed = TRUE)      
  )

# Print out the map
ny_map

```

Analysis:

With this map, it is easily seen where there is the most common location for
the incidents to occur. With a significant grouping of the incidents in
locations such as Brooklyn, The Bronx and even to the south East of Queens.
Which we can fact check by looking at the summary and pie chart
provided by the code below.The data points have also been color coded to see the
amount of incidents committed by race and in what areas they are occurring the
most often. It is easily noted that the there is the most significant amount of
incidents involving black perpetrators all through the different areas,
but with there being a significant amount of incidents involving Hispanic
perpetrators in the the Bronx. A following question would be what the ethnic 
population of those areas are as well as a follow up analysis and map could
be done one the victims races in those areas as well.

```{r incidents_per_boro_table}

incidents_per_boro <- labeled_data %>% 
  group_by(BORO) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))

incidents_per_boro

```

```{r incident_per_boro_pie_chart}

ggplot(incidents_per_boro, aes(x = "", y = count, fill = BORO)) +
  geom_bar(stat = "identity") +
  coord_polar("y", start = 0) +  # Convert to polar coordinates
  labs(title = "Incidents Per Boro", fill = "Categories") +  # Add title and legend
  theme_void() +  # Remove axes and background
  theme(plot.title = element_text(hjust = 0.5))

```



```{r density_graph}

# Create density plot showing the amount of incidents throughout the day
density_plot <- ggplot(cleaned_data, aes(x = OCCUR_TIME, fill = PERP_SEX)) +
  geom_density(alpha = 0.7) +
  labs(title = "Density Plot of Incidents Throughout the Day",
       x = "Time of Day",
       y = "Density") +
  theme_minimal() +
  scale_x_time(breaks = hms::hms(hours = seq(0, 24, by = 4)),
               labels = function(x) {
                 sprintf("%02d:%02d", hour(x), minute(x))
               }) +
  theme(plot.title = element_text(hjust = 0.5))

# Print the plot
print(density_plot)
```

Analysis:

This density plot clearly demonstrates the most likely time of day that an
incident would occur. Those being, as one might expect, very early in the
morning and very late into the night with peak activity being likely around 2am
and around 10pm. However, this data could likely be different depending on the
time of the year which would need further investigation. Additionally, the 
chart shows the difference, however ultimately slight, between the likelihood of
an incident occurring throughout the day by a Male or Female perpetrator.

```{r yearly_incident_activity_1}
# Extract month and year and make integers
labeled_data$month <- month(labeled_data$OCCUR_DATE) %>% 
  as.integer(labeled_data$month)
labeled_data$year <- year(labeled_data$OCCUR_DATE) %>% 
  as.integer(labeled_data$year)

# Create new start and end dates for new filtering
start_date2 <- as.Date("2018-01-01")
end_date2 <- as.Date("2022-12-31")

# Identify counts
yearly_activity <- labeled_data %>%
  filter(OCCUR_DATE >= start_date2 & OCCUR_DATE <= end_date2) %>% 
  group_by(month, year) %>%
  summarise(count = n())

# Create a variable with month names for x-axis labeling
custom_month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec") 

# Make the month a factor with the labels as the month names
yearly_activity$month <- factor(yearly_activity$month, levels = 1:12, labels = custom_month_names)

# Plot the yearly activity
ggplot(yearly_activity, aes(x = month, y = count, fill = factor(year))) +
  geom_bar(stat = "identity", position = "dodge") + # side by side bar chart
  labs(x = "Month", y = "Number of Incidents", fill = "Year") +
  scale_fill_discrete(name = "Year") + # filling in legend color by year
  theme_minimal() +
  ggtitle("Number of Incidents by Month and Year") +
  theme(plot.title = element_text(hjust = 0.5))


```


```{r yearly_incident_activity_2}

# Second plot for a different view
ggplot(yearly_activity, aes(x = month, y = count, color = factor(year), group = year)) +
  geom_point() +
  geom_line() +
  labs(x = "Month", y = "Number of Incidents", color = "Year") +
  scale_color_discrete(name = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of Incidents by Month and Year") +
  theme(plot.title = element_text(hjust = 0.5))

```

Analysis:

The last two charts show the trends of incidents both across multiple years as
well as throughout the months of the year.The bar chart depicts how the
incidents were much fewer in 2018 and 2019 and have steadily grown over the four years.
It is worth noting that COVID-19 had a great influence in the world during 2019
so this could also be possible of being influenced by COVID-19. The second chart
shows a more in depth look at the change of incident occurrences throughout the
year. Even with the fewer incidents in 208 and 2019, there are still more
significant spikes in occurrences during the warmer months and peaking in May,
June and July and then tapering off around September and October.



Conclusion:

In this report, we analyzed some of the variables involved in NYC shooting
incidents. We looked at map that demonstrated the most active locations for
these incidents to occur as well as the race of those who caused the incidents
in these locations.We could see by looking at the data that the most likely time
of day for these incidents to occur, appear to be around 2am and 10pm with a
slight variance in the time whether male or female perpetrator. In the last
two charts we observed that the number of incidents per year had increased and
that throughout the year the most active time for the occurrences is in the
warmer months.